Parameter Recovery Validation

### 2025-06-16 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): combine sections on dimensionless timescale
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

### Biological Interpretation of Timescale Prior Hyperparameters

The process duration $T_M$ determines the temporal scope of observable dynamics and requires careful hyperparameter selection based on biological expectations and experimental context.

#### **Dynamical Range and Steady-State Achievement**

@@

- $T_M = 1$: Observes 63% approach to steady state
- $T_M = 3$: Observes 95% approach to steady state
- $T_M = 5$: Observes >99% approach to steady state

@@

- **Biological assumption**: Process likely spans ~5 characteristic timescales
- **Dynamical expectation**: Will observe near-complete approach to steady state
- **Data burden**: Data must provide evidence if process is shorter than expected
- **Use case**: When biological knowledge suggests complete developmental programs

@@

- **Biological assumption**: Uncertain about process duration, wide range possible
- **Dynamical expectation**: Could observe early transients or complete dynamics
- **Data burden**: Data determines whether short or long timescales are needed
- **Use case**: When process duration is unknown, exploratory analysis

@@

- **Biological assumption**: Most cellular processes are rapid (early dynamics only)
- **Dynamical expectation**: Primarily observe transient phase (63% approach)
- **Data burden**: Data must strongly support longer timescales if present
- **Use case**: When expecting rapid responses, stress conditions, acute perturbations

@@

**Concrete examples for different experimental contexts:**
- **Conservative (Gamma(5,1))**: E[$T_M$] = 5 → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
- **Skeptical (Gamma(1,1))**: E[$T_M$] = 1 → $t^*_{on,i} \sim \text{Normal}(0.3, \sigma_{on}^2)$
- **Permissive (Gamma(1,0.2))**: E[$T_M$] = 5 → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$

@@

### Dimensionless Framework and Biological Timescale Connection

The dimensionless parameterization requires careful interpretation to connect mathematical parameters to biological reality. This section clarifies the relationship between dimensionless temporal parameters and real biological timescales.

**Biological interpretation examples:**

- $T_M = 1$: Process duration equals one characteristic timescale
- $T_M = 5$: Process duration equals five characteristic timescales
- $T_M = 0.5$: Process duration is half a characteristic timescale

**Steady-state achievement**: The fraction of steady-state dynamics observable is $1 - e^{-T_M}$:
- $T_M = 1$: Observes 63% approach to steady state
- $T_M = 3$: Observes 95% approach to steady state
- $T_M = 5$: Observes >99% approach to steady state

#### **Hyperparameter Choice as Biological Prior Knowledge**

**Conservative Prior: $T_M \sim \text{Gamma}(5.0, 1.0)$**

- $\mathbb{E}[T_M] = 5$, $\text{Var}[T_M] = 5$
- **Biological assumption**: "I expect this process to span approximately 5 characteristic timescales, allowing observation of near-complete dynamics"
- **Implications**: If experimental duration is 20 hours, assumes characteristic timescale ≈ 4 hours
- **Expectation**: Observes 99% approach to steady state ($1 - e^{-5} \approx 0.99$)
- **Data burden**: Data must provide strong evidence if process is much shorter or longer

**Permissive Prior: $T_M \sim \text{Gamma}(1.0, 0.2)$**

- $\mathbb{E}[T_M] = 5$, $\text{Var}[T_M] = 25$
- **Biological assumption**: "I'm uncertain about process duration - it could range from rapid (1 timescale) to very slow (15+ timescales)"
- **Implications**: Wide uncertainty allows data to determine appropriate timescale
- **Flexibility**: Can accommodate both early transient dynamics and complete processes
- **Data burden**: Minimal prior bias on process duration

**Skeptical Prior: $T_M \sim \text{Gamma}(1.0, 1.0)$**

- $\mathbb{E}[T_M] = 1$, $\text{Var}[T_M] = 1$
- **Biological assumption**: "I expect rapid cellular responses, likely observing only early dynamics"
- **Implications**: Assumes observation of primarily transient phase (63% approach: $1 - e^{-1} \approx 0.63$)
- **Data burden**: Data must strongly support longer timescales if present
- **Use cases**: Stress responses, drug treatments, acute perturbations

**Priors for other gene-specific parameters:**
\begin{align}
    U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)
\end{align}

**Cell-specific capture efficiency:**
\begin{align}
    \lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)
\end{align}

#### **Model Fitting and Timescale Inference**

The dimensionless framework creates a bidirectional relationship between process duration and characteristic rates:

**Forward (simulation):**

1. Choose biological timescales (e.g., β = splicing rate)
2. Set process duration in real time (e.g., 24 hours)
3. Compute $T_M = t_{\text{process}} \times β$
4. Generate dimensionless dynamics

**Inverse (inference from data):**

1. Observe expression data over known duration
2. Infer $T_M$ from temporal patterns
3. Estimate characteristic rates: $\hat{β} = T_M/t_{\text{experiment}}$
4. Convert all dimensionless parameters to real time

**Practical Example:**

**Experimental setup:** 48-hour differentiation time course

**Model fitting results:**

- Posterior: $T_M \approx 6.0 \pm 0.5$
- Inferred characteristic timescale: $48/6 = 8$ hours
- Gene with $t^*_{on,i} = 2.0$: Activates at $2.0 \times 8 = 16$ hours
- Gene with $\delta^*_i = 1.5$: Active for $1.5 \times 8 = 12$ hours

>>>>

### Biological Interpretation of Timescale Prior Hyperparameters and Dimensionless Framework

The dimensionless parameterization requires careful interpretation to connect mathematical parameters to biological reality. This section clarifies the relationship between dimensionless temporal parameters and real biological timescales, with particular focus on how hyperparameter choices encode biological prior knowledge.

**Dynamical Range and Steady-State Achievement**

@@

- $T_M = 1$: Process duration equals one characteristic timescale, observes 63% approach to steady state
- $T_M = 3$: Process duration equals three characteristic timescales, observes 95% approach to steady state
- $T_M = 5$: Process duration equals five characteristic timescales, observes >99% approach to steady state

@@

- **Biological assumption**: "I expect this process to span approximately 5 characteristic timescales, allowing observation of near-complete dynamics"
- **Dynamical expectation**: Will observe near-complete approach to steady state (99%: $1 - e^{-5} \approx 0.99$)
- **Data burden**: Data must provide strong evidence if process is much shorter or longer
- **Use case**: When biological knowledge suggests complete developmental programs

@@

- **Biological assumption**: "I'm uncertain about process duration - it could range from rapid (1 timescale) to very slow (15+ timescales)"
- **Dynamical expectation**: Could observe early transients or complete dynamics
- **Data burden**: Minimal prior bias on process duration, data determines whether short or long timescales are needed
- **Use case**: When process duration is unknown, exploratory analysis

@@

- **Biological assumption**: "I expect rapid cellular responses, likely observing only early dynamics"
- **Dynamical expectation**: Primarily observe transient phase (63% approach: $1 - e^{-1} \approx 0.63$)
- **Data burden**: Data must strongly support longer timescales if present
- **Use case**: When expecting rapid responses, stress conditions, acute perturbations

@@

**Concrete examples for different experimental contexts:**

- **Conservative (Gamma(5,1))**: E[$T_M$] = 5 → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
- **Skeptical (Gamma(1,1))**: E[$T_M$] = 1 → $t^*_{on,i} \sim \text{Normal}(0.3, \sigma_{on}^2)$
- **Permissive (Gamma(1,0.2))**: E[$T_M$] = 5 → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$

@@
